Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-13 paper-new/terms.tex
changes to term syntax + semantics

In the following section, we will extend the denotational semantics to accommodate
the type and term precision orderings.

@@

\subsection{Removing Casts as Primitives}

@@

\section{Term Semantics}\label{sec:term-semantics}

\subsection{Domain-Theoretic Constructions}\label{sec:domain-theory}

In this section, we discuss the fundamental objects of the model into which we will embed
the gradual lambda calculus.
It is important to remember that the constructions in this section are entirely
independent of the syntax described in the previous section; the notions defined 
here exist in their own right as purely mathematical constructs.
In Section \ref{sec:term-interpretation}, we will link the syntax and semantics
via a semantic interpretation function.

@@

% TODO explain better
As it turns out, this definition is not quite correct; let us try to understand why.
The price we pay for such a simple solution to the above equation is that we have
introduced a notion of ``time'', i.e., we must now consider the stepping behavior of terms.
% Therefore, the semantics of terms will need to allow for terms that potentially
% do not terminate. This is accomplished using an extension of the guarded lift monad,
% which we describe in the next section.
Thus, in the equation for the semantics of $\dyn$, the result of the function should be
a computation that may return a value, error, \emph{or} take an observable step.
We model such computations using an extension of the so-called guarded lift monad
\cite{mogelberg-paviotti2016} which we describe in the next section.
We will then discuss the correct definition of the semantics of the dynamic type.

@@

% TODO ensure the previous section flows into this one
When thinking about how to model gradually-typed programs where we track the steps they take,
we should consider their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
at run-time because of a type error. In addition to this, a program may ``think'',
i.e., take a step of computation. If a program thinks forever, then it never returns a value,
so we can think of the idea of thinking as a way of intensionally modelling \emph{partiality}.

With this in mind, we can describe a semantic object that models these behaviors: a monad
for embedding computations that has cases for failure and ``thinking''.
Previous work has studied such a construct in the setting of the latter, called the lift
monad \cite{mogelberg-paviotti2016}; here, we augment it with the additional effect of failure.

@@

An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
(2) raise an error and stop (via $\mho$), or (3) think for a step (via $\theta$).
%
Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
that runs forever and never returns a value.

@@

We can now give a semantics to the gradual lambda calculus we defined
in Section \ref{sec:step-sensitive-lc}.
%
Much of the semantics is similar to a normal call-by-value denotational semantics;
we highlight the differences.
We will interpret types as sets, and terms as functions.
Contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$
will be interpreted as the product $\sem{A_1} \times \cdots \times \sem{A_n}$.

The semantics of the dynamic type $\dyn$ is the set $\Dyn$ introduced in Section
\ref{sec:predomains}.
%
The interpretation of a value $\vhasty {\Gamma} V A$ will be a monotone function from
$\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\phasty {\Gamma} M {{A}}$ will be interpreted
as a monotone function from $\sem{\Gamma}$ to $\li \sem{A}$.

@@

The interpretation of $\lda{x}{M}$ works as follows. Recall by the typing rule for
lambda that $\phasty {\Gamma, x : A_i} M {{A_o}}$, so the interpretation of $M$
has type $(\sem{\Gamma} \times \sem{A_i})$ to $\li \sem{A_o}$.
The interpretation of lambda is thus a function (in the ambient type theory) that takes
a value $a$ representing the argument and applies it (along with $\gamma$) as argument to
the interpretation of $M$.
%
The interpretation of bind and of application both make use the monadic extend function on $\li A$.
%
The interpretation of case-nat and case-arrow is simply a case inspection on the
interpretation of the scrutinee, which has type $\Dyn$.

@@

\noindent Types:
\begin{align*}
  \sem{\nat} &= \Nat \\
  \sem{\dyn} &= \Dyn \\
  \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
\end{align*}

@@

% TODO check these, especially the semantics of bind, case-nat, and case-arr
% with respect to their context argument
\noindent Values and terms:
\begin{align*}
  \sem{\zro}         &= \lambda \gamma . 0 \\
  \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
  \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
  \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
  \sem{\injnat{V_n}} &= \lambda \gamma . \inl\, (\sem{V_n}\, \gamma) \\
  \sem{\injarr{V_f}} &= \lambda \gamma . \inr\, (\sem{V_f}\, \gamma) \\[2ex]
  \sem{\nxt\, V}     &= \lambda \gamma . \nxt (\sem{V}\, \gamma) \\
  \sem{\theta}       &= \lambda \gamma . \theta \\
%
  \sem{\err_B}         &= \lambda \delta . \mho \\
  \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
  \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
  \sem{V_f\, V_x}      &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
  \sem{\casedyn{V}{n}{M_{nat}}{\tilde{f}}{M_{fun}}} &=
    \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
    &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
    &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
\end{align*}

>>>>

In Section \ref{sec:gtlc-precision}, we will discuss how to extend the denotational
semantics to accommodate the type and term precision orderings.

@@

\begin{comment}
\subsection{Removing Casts as Primitives}

@@

\end{comment}

% \section{Term Semantics}\label{sec:term-semantics}

\subsection{Semantic Constructions}\label{sec:domain-theory}

In this section, we discuss the fundamental objects of the model into which we will embed
the terms of the gradually-typed lambda calculus.
It is important to remember that the constructions in this section are entirely
independent of the syntax described in the previous section; the notions defined 
here exist in their own right as purely mathematical constructs.
In Section \ref{sec:term-interpretation}, we will link the syntax and semantics
via a semantic interpretation function.

@@

% TODO explain better
As it turns out, this definition is not quite correct, as it doesn't provide a way to
model functions that are potentially non-terminating.
Another way to think about this is that by using guarded recursion to solve the
equation for the dynamic type, the solution to the equation involves a notion of
``time" or ``steps".
So, in addition to returning a value or erroring, programs may now take one or
more observable steps of computation, and this possibility must be reflected in
in the equation for the dynamic type.

% Therefore, the semantics of terms will need to allow for terms that potentially
% do not terminate. This is accomplished using an extension of the guarded lift monad,
% which we describe in the next section.
More specifically, in the equation for the semantics of $\dyn$, the result of the
function should be a computation that may return a value, error, \emph{or} take an observable step.
We model such computations using an extension of the so-called guarded lift monad
\cite{mogelberg-paviotti2016} which we describe in the next section.
We will then use this to give the correct definition of the semantics of the dynamic type.

@@

% TODO ensure the previous section flows into this one
When thinking about how to model gradually-typed programs where we track the steps they take,
we should consider their possible behaviors. On the one hand, we have \emph{failure}: a program may fail
at run-time because of a type error. In addition to this, a program may take one or more steps of computation.
If a program steps forever, then it never returns a value,
so we can think of the idea of stepping as a way of intensionally modelling \emph{partiality}.

With this in mind, we can describe a semantic object that models these behaviors: a monad
for embedding computations that has cases for failure and ``stepping''.
Previous work has studied such a construct in the setting of the latter, called the lift
monad \cite{mogelberg-paviotti2016}; here, we augment it with the additional effect of failure.

@@

An element of $\li A$ should be viewed as a computation that can either (1) return a value (via $\eta$),
(2) raise an error and stop (via $\mho$), or (3) take a step (via $\theta$).
%
Notice there is a computation $\fix \theta$ of type $\li A$. This represents a computation
that runs forever and never returns a value.

@@

We can now give a semantics to the \emph{terms} of the gradual lambda calculus we defined
above. The full definition is given in Figure \ref{fig:term-semantics}.
%
Much of the semantics is similar to a normal call-by-value denotational semantics:
We will interpret types as sets, and terms as functions.
Contexts $\Gamma = x_1 \colon A_1, \dots, x_n \colon A_n$
will be interpreted as the product $\sem{A_1} \times \cdots \times \sem{A_n}$.

The semantics of the dynamic type $\dyn$ is the set $\Dyn$ introduced in Section
\ref{sec:predomains}.
%
The interpretation of a value $\vhasty {\Gamma} V A$ will be a function from
$\sem{\Gamma}$ to $\sem{A}$. Likewise, a term $\phasty {\Gamma} M {{A}}$ will be interpreted
as a function from $\sem{\Gamma}$ to $\li \sem{A}$.

@@

The interpretation of $\lda{x}{M}$ works as follows. Recall by the typing rule for
lambda that $\phasty {\Gamma, x : A_i} M {{A_o}}$, so the interpretation of $M$
has type $(\sem{\Gamma} \times \sem{A_i})$ to $\li \sem{A_o}$.
The interpretation of lambda is thus a function (in the ambient type theory) that takes
a value $a$ representing the argument and applies it (along with $\gamma$) as argument to
the interpretation of $M$.
%
The interpretation of bind makes use the monadic extend function on $\li A$.
%
The interpretation of case-nat and case-arrow is simply a case inspection on the
interpretation of the scrutinee, which has type $\Dyn$.

@@

\begin{figure*}
  \noindent Types:
  \begin{align*}
    \sem{\nat} &= \Nat \\
    \sem{\dyn} &= \Dyn \\
    \sem{A \ra A'} &= \sem{A} \To \li \sem{A'} \\
  \end{align*}

@@

  % TODO check these, especially the semantics of bind, case-nat, and case-arr
  % with respect to their context argument
  \noindent Values and terms:
  \begin{align*}
    \sem{\zro}         &= \lambda \gamma . 0 \\
    \sem{\suc\, V}     &= \lambda \gamma . (\sem{V}\, \gamma) + 1 \\
    \sem{x \in \Gamma} &= \lambda \gamma . \gamma(x) \\
    \sem{\lda{x}{M}}   &= \lambda \gamma . \lambda a . \sem{M}\, (*,\, (\gamma , a))  \\
    \sem{\injnat{V_n}} &= \lambda \gamma . \inl\, (\sem{V_n}\, \gamma) \\
    \sem{\injarr{V_f}} &= \lambda \gamma . \inr\, (\sem{V_f}\, \gamma) \\[2ex]
    % \sem{\nxt\, V}     &= \lambda \gamma . \nxt (\sem{V}\, \gamma) \\
    % \sem{\theta}       &= \lambda \gamma . \theta \\
  %
    \sem{\err_B}         &= \lambda \delta . \mho \\
    \sem{\ret\, V}       &= \lambda \gamma . \eta\, \sem{V} \\
    \sem{\bind{x}{M}{N}} &= \lambda \delta . \ext {(\lambda x . \sem{N}\, (\delta, x))} {\sem{M}\, \delta} \\
    \sem{V_f\, V_x}      &= \lambda \gamma . {({(\sem{V_f}\, \gamma)} \, {(\sem{V_x}\, \gamma)})} \\
    \sem{\casedyn{V}{n}{M_{nat}}{\tilde{f}}{M_{fun}}} &=
      \lambda \delta . \text{case $(\sem{V}\, \delta)$ of} \\ 
      &\quad\quad\quad\quad \alt \inl(n) \to \sem{M_{nat}}(n) \\
      &\quad\quad\quad\quad \alt \inr(\tilde{f}) \to \sem{M_{fun}}(\tilde{f})
  \end{align*}

  \caption{Term semantics for the gradually-typed lambda calculus.}
  \label{fig:term-semantics}
\end{figure*}
